Deep Reinforcement Learning for Conservation Decisions

### 2021-06-13 appendices/appendix-A-definitions.Rmd
Most significant edit to the RL tables; diettrich citation added also

Reinforcement learning is a huge, constantly evolving field, and we've only touched on a small sliver of RL in our study.
In this appendix, we briefly introduced some of the major design decisions that are made with model-free deep RL algorithms, but there are many more decisions that are made with state of the art model-free agents.
In Table 1, we collect some of the more commonly used deep RL algorithms.
Readers who are interested in learning more about the current state of deep RL should read the literature referenced in Table 1; @DQN, @sac and @ppo are good places to start. 
It is important to note that while we have focused on gradient-based, model-free methods, there has also been promising recent development with model-based and gradient-free techniques.
See @mbpo for a recent example of a model-based deep RL algorithm and @salimans_evolution_2017 as a recent example of a gradient-free approach.
A benefit of model-free deep RL over these other methods is that there has been more of a general focus on model-free deep RL, so there are widely available software packages like Stable Baselines and TensorFlow Agents which implement model-free deep RL agents.
However, it is important to keep in mind though that deep RL has only become established in the last decade, and with the large amount of open research questions, deep RL may evolve significantly in the near future.

\begin{table}[H]
\resizebox{\textwidth}{!}{\begin{tabular}{l l l l l}
\hline
Abbreviation & Algorithm Name                                                   & Model       & Policy     & Method \\
\hline
MBPO         & Model-based Policy Optimization (Janner et al. 2019)              & Model-based & On-policy  & Policy Gradient \\
MCTS         & Monte Carlo Tree Search (Sutton and Barto 2018)                   & Model-based & Either     & Any\\
A2C          & Advantage Actor Critic  (Mnih et al. 2016)                        & Model-free  & On-policy  & Actor-critic\\
A3C          & Asynchronous A2C  (Babaeizadeh et al. 2017)                       & Model-free  & On-policy  & Actor-critic\\
TRPO         & Trust Region Policy Optimization  (Schulman, Levine,et al. 2017)  & Model-free  & On-policy  & Policy Gradient\\
PPO          & Proximal Policy Optimization  (Schulman, Wolski, et al. 2017)     & Model-free  & On-policy  & Actor-critic\\
DQN          & Deep Q Networks   (Mnih et al. 2015)                              & Model-free  & Off-policy & Value-Based\\
DDPG         & Deep Deterministic Policy Gradient   (Lillicrap et al. 2019)      & Model-free  & Off-policy & Actor-critic\\
TD3          & Twin Delayed DDPG  (Fujimoto, Hoof, and Meger 2018)               & Model-free  & Off-policy & Actor-critic\\
SAC          & Soft Actor Critic   (Haarnoja et al. 2018)                        & Model-free  & Off-policy & Actor-critic\\
IMPALA       & Importance Weighted Actor Learner (Espeholt et al. 2018)          & Model-free  & Off-policy & Actor-critic\\
\end{tabular}}
\caption{Survey of common deep RL algorithms. }
\end{table}

>>>>

Reinforcement learning is a huge, constantly evolving field, and we've only touched on a small sliver of RL in our study.
In this appendix, we briefly introduced some of the major design decisions that are made with model-free deep RL algorithms, but there are many more decisions that are made with state of the art model-free agents.
In Table 1, we collect some of the more commonly used model-free deep RL algorithms.
Readers who are interested in learning more about the current state of deep RL should read the literature referenced in Table 1; @DQN, @sac and @ppo are good places to start. 
It is important to note that while we have focused on gradient-based, model-free methods, there has also been promising recent development with model-based and gradient-free techniques.
See @mbpo and @weber_imagination-augmented_2018 for a recent examples of model-based deep RL algorithms, and @salimans_evolution_2017 as a recent example of a gradient-free approach.
A benefit of model-free deep RL over these other methods is that there has been more of a general focus on model-free deep RL, so there are widely available software packages like Stable Baselines and TensorFlow Agents which implement model-free deep RL agents.
However, it is important to keep in mind though that deep RL has only become established in the last decade, and with the large amount of open research questions, deep RL may evolve significantly in the near future.

\begin{table}[H]
\resizebox{\textwidth}{!}{\begin{tabular}{l l l l l}
\hline
Abbreviation & Algorithm Name                                                   & Model       & Policy     & Method \\
\hline
A2C          & Advantage Actor Critic  (Mnih et al. 2016)                        & Model-free  & On-policy  & Actor-critic\\
A3C          & Asynchronous A2C  (Babaeizadeh et al. 2017)                       & Model-free  & On-policy  & Actor-critic\\
TRPO         & Trust Region Policy Optimization  (Schulman, Levine,et al. 2017)  & Model-free  & On-policy  & Policy Gradient\\
PPO          & Proximal Policy Optimization  (Schulman, Wolski, et al. 2017)     & Model-free  & On-policy  & Actor-critic\\
DQN          & Deep Q Networks   (Mnih et al. 2015)                              & Model-free  & Off-policy & Value-Based\\
DDPG         & Deep Deterministic Policy Gradient   (Lillicrap et al. 2019)      & Model-free  & Off-policy & Actor-critic\\
TD3          & Twin Delayed DDPG  (Fujimoto, Hoof, and Meger 2018)               & Model-free  & Off-policy & Actor-critic\\
SAC          & Soft Actor Critic   (Haarnoja et al. 2018)                        & Model-free  & Off-policy & Actor-critic\\
IMPALA       & Importance Weighted Actor Learner (Espeholt et al. 2018)          & Model-free  & Off-policy & Actor-critic\\
\end{tabular}}
\caption{Survey of common model-free deep RL algorithms. }
\end{table}

### 2021-06-13 manuscript/manuscript.Rmd
Most significant edit to the RL tables; diettrich citation added also

---
title: Deep Reinforcement Learning for Conservation Decisions
authors:
  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: mlapeyro@berkeley.edu
  - name: Melissa S. Chapman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: mchapman@berkeley.edu
  - name: Kari E. A. Norman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: kari.norman@berkeley.edu
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu (corresponding author)
abstract: |
  Can machine learning help us make better decisions about a changing planet?
  In this paper, we illustrate and discuss the potential of a promising 
  corner of machine learning known as _reinforcement learning_ (RL) to help tackle the
  most challenging conservation decision problems. RL is uniquely
  well suited to conservation and global change challenges for three reasons: (1)
  RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
  (2) RL approaches do not require massive amounts of data, (3) RL approaches would 
  utilize rather than replace existing models, simulations, and the knowledge they contain.
  We provide a conceptual and technical introduction to RL and its relevance to ecological
  and conservation challenges, including examples of a problem in setting
  fisheries quotas and in managing ecological tipping points.
  Four appendices with annotated code provide a tangible introduction to researchers 
  looking to adopt, evaluate, or extend these approaches.

keywords:
  - Reinforcement Learning
  - Conservation
  - Machine Learning
  - Artificial Intelligence
  - Tipping points
bibliography: references.bib
nocite: |
    @A2C, @trpo, @ddpg, @TD3, @impala
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
output: 
  rticles::arxiv_article:
    keep_tex: true
  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

a short running title: **Deep Reinforcement Learning**

the type of article: **Method**

the number of words: **4,814**

the number of references: **62**

the number of figures, tables, and text boxes: **6** (five figures, 1 table)

@@

Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in these fields.
To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@openai_learning_2019; @alphazero; @alphastar].
Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
RL agents typically develop strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is common across ecology.
Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or climate change policy models [@dice] are already used to explore scenarios and inform ecosystem management.
Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013].
Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.

In this paper, we draw on examples from fisheries management and ecological tipping points to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
We demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
We include an extensive appendices with carefully annotated code which should allow readers to both reproduce and extend this analysis.

@@

An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of the environment's current *state* and an associated *reward* as output.
In contrast to classical approaches [@Marescot2013], there are few restrictions on what comprises a state or action. 
States and actions may be continuous or discrete, completely or partially observed, single or multidimensional.
The main focus of building an RL environment, however, is in the environment's transition dynamics and reward function.
The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and action.
This ability allows RL environments to model a broad range of decision making problems.
For example, we can set the transitions to be deterministic or stochastic.
We can also specify the reward function to be *sparse*, whereby a positive reward can only be received after a long sequence of actions, e.g. the end point in a maze.
In other environments, an agent may have to learn to forgo immediate rewards (or even accept an initial negative reward) in order to maximize the net discounted reward as we illustrate in examples here.

@@

\begin{table}
\begin{tabular}{l l l}
\hline
Abbreviation & Algorithm Name                                                    & Model \\
\hline
MCTS         & Monte Carlo Tree Search (Sutton and Barto 2018)                   & Model-based \\
DYNA         & Dyna-Q (Sutton and Barto 2018)                                    & Model-based \\
MBPO         & Model-based Policy Optimization (Janner et al. 2019)              & Model-based \\
DQN          & Deep Q Networks   (Mnih et al. 2015)                              & Model-free  \\
A2C          & Advantage Actor Critic  (Mnih et al. 2016)                        & Model-free  \\
A3C          & Asynchronous A2C  (Babaeizadeh et al. 2017)                       & Model-free  \\
TRPO         & Trust Region Policy Optimization  (Schulman, Levine,et al. 2017)  & Model-free  \\
PPO          & Proximal Policy Optimization  (Schulman, Wolski, et al. 2017)     & Model-free  \\
DDPG         & Deep Deterministic Policy Gradient   (Lillicrap et al. 2019)      & Model-free  \\
TD3          & Twin Delayed DDPG  (Fujimoto, Hoof, and Meger 2018)               & Model-free  \\
SAC          & Soft Actor Critic   (Haarnoja et al. 2018)                        & Model-free  \\
IMPALA       & Importance Weighted Actor Learner (Espeholt et al. 2018)          & Model-free  \\
\hline
\end{tabular}
\caption{Survey of common deep RL algorithms.}
\end{table}

@@

$$
  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi).
$$

Although there are RL-based methods for infinite horizon problems, i.e. when $H=\infty$, we will only present episodic or finite horizon POMDPs in this study.
In Appendix A, we will discuss in greater detail how deep RL algorithms attempt to optimize the RL objective.

@@

We provide two examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and the inherent challenges.
Annotated code for these examples may be found in Appendix B and at <https://github.com/boettiger-lab/rl-intro>.

@@

A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy [see Appendix B] further highlights the difference between these two approaches.
Under the steady-state strategy, the system remains poised too close to the tipping point: stochastic noise eventually drives most replicates across the threshold, where the steady-state strategy is too weak to bring them back once they collapse.
As replicate after replicate stochastically crashes, the mean state and mean reward bend increasingly downwards.
In contrast, the RL agent edges the system slightly farther away from the tipping point, decreasing but not eliminating the chance of a chance transition.
In the few replicates that experience a critical transition anyway, the RL agent usually responds with sufficient commitment to ensure their recovery [Appendix B].
Only 3 out of 100 replicates degrade far enough for the RL agent to give up the high cost of trying to rescue them.
The RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
Numerous kinks visible in the RL policy function also suggest that this solution is not yet optimal.
Further tuning of hyper-parameters, increased training, alterations or alternatives to the training algorithm would likely be able to further improve upon this performance.
Nevertheless, such quirks are likely to be common features of RL solutions -- long as they have minimal impact on realized rewards.

@@

Ecology holds many open problems for deep RL.
To extend the examples presented here to reflect greater biological complexity or more realistic decision scenarios, the transition, emission and/or reward functions of the environment can be modified.
We provide an initial library of example environments at <https://boettiger-lab.github.io/conservation-gym>.
Some environments in this library include a wildfire `gym` that poses the problem on wildfire suppression with a cellular automata model, an epidemic `gym` that examines timing of interventions to curb disease spread, as well as more complex variations of the fishing and conservation environments presented above.

>>>>

---
title: Deep Reinforcement Learning for Conservation Decisions
authors:
  - name: Marcus Lapeyrolerie
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Melissa S. Chapman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Kari E. A. Norman
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
  - name: Carl Boettiger
    department: Department of Environmental Science, Policy, and Management
    affiliation: University of California, Berkeley
    location: Berkeley, California
    email: cboettig@berkeley.edu
abstract: |
  Can machine learning help us make better decisions about a changing planet?
  In this paper, we illustrate and discuss the potential of a promising 
  corner of machine learning known as _reinforcement learning_ (RL) to help tackle the
  most challenging conservation decision problems. RL is uniquely
  well suited to conservation and global change challenges for three reasons: (1)
  RL explicitly focuses on designing an agent who _interacts_ with an environment which is dynamic and uncertain,
  (2) RL approaches do not require massive amounts of data, (3) RL approaches would 
  utilize rather than replace existing models, simulations, and the knowledge they contain.
  We provide a conceptual and technical introduction to RL and its relevance to ecological
  and conservation challenges, including examples of a problem in setting
  fisheries quotas and in managing ecological tipping points.
  Four appendices with annotated code provide a tangible introduction to researchers 
  looking to adopt, evaluate, or extend these approaches.

keywords:
  - Reinforcement Learning
  - Conservation
  - Machine Learning
  - Artificial Intelligence
  - Tipping points
bibliography: references.bib
nocite: |
    @A2C, @trpo, @ddpg, @TD3, @impala, @lenz_deepmpc_2015, @weber_imagination-augmented_2018
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
output: 
  rticles::arxiv_article:
    keep_tex: true
  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

@@

Despite relevance to decision making under uncertainty that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in these fields.
To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@openai_learning_2019; @alphazero; @alphastar].
Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
RL agents typically develop strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is common across ecology.
Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or climate change policy models [@dice] are already used to explore scenarios and inform ecosystem management.
Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods [@Possingham2006; @Marescot2013; @ferrer-mestres_k-n-momdps_2021].
Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.

In this paper, we draw on examples from fisheries management and ecological tipping points to illustrate how deep RL techniques can successfully discover optimal solutions to previously solved management scenarios and discover highly effective solutions to unsolved problems.
We demonstrate that RL-based approaches are capable but by no means a magic bullet: reasonable solutions require careful design of training environments, choice of RL algorithms, tuning and evaluation, as well as substantial computational power.
Our examples are intentionally simple, aiming to provide a clear template for understanding that could be easily extended to cover more realistic conditions. 
We include an extensive appendix with carefully annotated code which should allow readers to both reproduce and extend this analysis.

@@

An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of the environment's current *state* and an associated *reward* as output.
In contrast to classical approaches [@Marescot2013], there are few restrictions on what comprises a state or action. 
States and actions may be continuous or discrete, completely or partially observed, single or multidimensional.
The main focus of building an RL environment, however, is on the environment's transition dynamics and reward function.
The designer of the environment can make the environment follow any transition and reward function provided that both are functions of the current state and action.
This ability allows RL environments to model a broad range of decision making problems.
For example, we can set the transitions to be deterministic or stochastic.
We can also specify the reward function to be *sparse*, whereby a positive reward can only be received after a long sequence of actions, e.g. the end point in a maze.
In other environments, an agent may have to learn to forgo immediate rewards (or even accept an initial negative reward) in order to maximize the net discounted reward as we illustrate in examples here.

@@

\begin{table}
\begin{tabular}{l l l}
\hline
Abbreviation & Algorithm Name                                                    & Model \\
\hline
D-MPC        & Deep Model-Predictive Control (Lenz, Knepper, and Saxena 2015)    & Model-based \\
I2A          & Imagination-Augmented Agents (Weber et al. 2018)                  & Model-based \\
MBPO         & Model-based Policy Optimization (Janner et al. 2019)              & Model-based \\
DQN          & Deep Q Networks   (Mnih et al. 2015)                              & Model-free  \\
A2C          & Advantage Actor Critic  (Mnih et al. 2016)                        & Model-free  \\
A3C          & Asynchronous A2C  (Babaeizadeh et al. 2017)                       & Model-free  \\
TRPO         & Trust Region Policy Optimization  (Schulman, Levine,et al. 2017)  & Model-free  \\
PPO          & Proximal Policy Optimization  (Schulman, Wolski, et al. 2017)     & Model-free  \\
DDPG         & Deep Deterministic Policy Gradient   (Lillicrap et al. 2019)      & Model-free  \\
TD3          & Twin Delayed DDPG  (Fujimoto, Hoof, and Meger 2018)               & Model-free  \\
SAC          & Soft Actor Critic   (Haarnoja et al. 2018)                        & Model-free  \\
IMPALA       & Importance Weighted Actor Learner (Espeholt et al. 2018)          & Model-free  \\
\hline
\end{tabular}
\caption{Survey of common deep RL algorithms.}
\end{table}

@@

$$
  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H-1} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi).
$$

Although there are RL-based methods for infinite horizon problems, i.e. when $H=\infty$, we will only present episodic or finite horizon POMDPs in this study.
In appendix A, we will discuss in greater detail how deep RL algorithms attempt to optimize the RL objective.

@@

We provide two examples that illustrate the application and potential of deep RL to ecological and conservation problems, highlighting both the potential and the inherent challenges.

@@

A close examination of the trajectories of individual simulations which cross the tipping point under either management strategy [see Appendix B] further highlights the difference between these two approaches.
Under the steady-state strategy, the system remains poised too close to the tipping point: stochastic noise eventually drives most replicates across the threshold, where the steady-state strategy is too weak to bring them back once they collapse.
As replicate after replicate stochastically crashes, the mean state and mean reward bend increasingly downwards.
In contrast, the RL agent edges the system slightly farther away from the tipping point, decreasing but not eliminating the chance of a chance transition.
In the few replicates that experience a critical transition anyway, the RL agent usually responds with sufficient commitment to ensure their recovery [Appendix B].
Only 3 out of 100 replicates degrade far enough for the RL agent to give up the high cost of trying to rescue them.
The RL agent's use of a more dynamic strategy out-preforms the steady-state strategy.
Numerous kinks visible in the RL policy function also suggest that this solution is not yet optimal.
Such quirks are likely to be common features of RL solutions -- long as they have minimal impact on realized rewards.
Further tuning of hyper-parameters, increased training, alterations or alternatives to the training algorithm would likely be able to further improve upon this performance.

@@

Ecology holds many open problems for deep RL.
To extend the examples presented here to reflect greater biological complexity or more realistic decision scenarios, the transition, emission and/or reward functions of the environment can be modified.
We provide an initial library of example environments at <https://boettiger-lab.github.io/conservation-gym>.
Some environments in this library include a wildfire `gym` that poses the problem of wildfire suppression with a cellular automata model, an epidemic `gym` that examines timing of interventions to curb disease spread, as well as more complex variations of the fishing and conservation environments presented above.
